% -------------------------------------------------------
% Abstract (English, unnumbered)
% -------------------------------------------------------

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

\begin{otherlanguage}{english}
Crystal sedimentation in viscous magmatic fluids is a fundamental geodynamic
process that drives chemical differentiation and controls the internal structure
of magma reservoirs. Accurate simulation of multi-crystal settling requires the
simultaneous resolution of microscale boundary layers around individual crystals
and long-range hydrodynamic interactions spanning the full computational domain.
This multi-scale character, combined with the near-quadratic scaling of
interaction costs with particle number, renders systematic parameter studies
with classical solvers computationally prohibitive.

This thesis investigates whether data-driven surrogate models can effectively
replace high-fidelity Stokes solvers for predicting flow fields in multi-crystal
sedimentation scenarios. Two neural network architectures are systematically
compared on identical data and evaluation metrics: the U-Net, an encoder--decoder
model with skip connections that extracts spatially hierarchical features, and
the Fourier Neural Operator (FNO), which operates in the spectral domain and
provides a global receptive field per layer. Both models are trained to predict
the stream function $\psi$ from a five-channel geometric input encoding consisting
of a crystal mask, a signed distance field, the radius of the nearest crystal,
and normalized spatial coordinates. Predicting $\psi$ rather
than velocity components directly enforces incompressibility by construction,
eliminating a common source of systematic constraint violations in purely
data-driven velocity surrogates.

Training data are generated using the \textsc{LaMEM} geodynamic finite-difference
code, providing steady-state incompressible Stokes solutions for configurations
of $1$ to $N_{\max}$ randomly placed rigid circular crystals on a fixed
$256 \times 256$ grid. Four experiments are conducted to compare architectural
performance and out-of-distribution (OOD) generalization. In the single-crystal
baseline (Experiment~1), the FNO achieves mean relative $L_2$ errors of
$1.7$--$2.3\%$ on~$\psi$, compared to $5.5$--$6.5\%$ for the U-Net, while
exhibiting more stable training dynamics. In multi-crystal settings
(Experiments~2 and~3, trained on up to $N_{\max} = 10$ and $25$ crystals,
respectively), the FNO achieves in-distribution errors of $4$--$8\%$,
maintaining a roughly five-fold advantage over the U-Net ($24$--$30\%$). OOD
generalization improves substantially when the training distribution is widened,
confirming that training distribution width is the primary factor determining
the generalization range. Experiment~4 reveals a reversal of the architectural
ranking for crystal-size generalization: for crystals larger than the training
distribution, the FNO degrades severely ($71.3 \pm 4.8\%$ on~$\psi$) while the
U-Net shows substantially lower errors ($44.9 \pm 8.2\%$), attributable to the
FNO's fixed spectral truncation at $k_{\max} = 16$ Fourier modes. A
variable-radius FNO trained on three discrete crystal sizes reduces this
degradation to $10.4\%$ through explicit size conditioning via the radius input
channel, demonstrating that the spectral limitation can be overcome without
architectural modifications.

Gradient-derived velocity quantities prove consistently harder to
reproduce than $\psi$ itself, as numerical differentiation amplifies
high-frequency residuals in the predicted stream function, disproportionately
increasing the velocity error. The stream-function representation guarantees
exact incompressibility for all predictions, confirmed by divergence RMS values
on the order of $10^{-26}$.

The controlled comparison of U-Net and FNO architectures on this benchmark
quantifies the trade-off between spatially local multiscale processing and global
spectral representations for surrogate modeling of elliptic, multi-inclusion
Stokes flow. The results provide practical guidelines for input encoding, output
representation, and architecture selection in geophysical sedimentation
applications.
\end{otherlanguage}
