\chapter{Theory and Physical Background}
\label{ch:theory}

This chapter summarizes the physical, mathematical, and architectural
foundations underlying the surrogate modeling approaches developed in this
thesis. The first part covers low-Reynolds-number fluid mechanics and the
stream-function formulation, which motivate the choice of learning target. The
second part introduces the two neural network architectures compared in this
work—the U-Net and the Fourier Neural Operator (FNO)—and discusses their
structural properties in the context of flow-field prediction.

% ==========================================================
\section{Incompressible Stokes Flow at Low Reynolds Number}
\label{sec:stokes_flow}
% ==========================================================

Crystal settling in magmatic systems typically occurs at very low Reynolds numbers, where viscous forces dominate over inertial effects \parencite{martin_crystal_1988,ladd_numerical_nodate}. In this creeping-flow regime, the governing equations reduce to the incompressible Stokes equations
\begin{equation}
\mathbf{0} = -\nabla p + \mu \nabla^2 \mathbf{u} + \rho \mathbf{g},
\qquad
\nabla \cdot \mathbf{u} = 0,
\end{equation}
where $\mathbf{u}$ denotes the velocity field, $p$ the pressure, $\mu$ the
dynamic viscosity, $\rho$ the density, and $\mathbf{g}$ the gravitational
acceleration.

A key property of Stokes flow is its linearity with respect to the
velocity field for fixed material parameters and boundary conditions. This
means that hydrodynamic interactions between rigid inclusions are long-ranged
and additive: the disturbance induced by one particle influences the
flow field throughout the domain. These long-range interactions form the
physical basis for collective sedimentation phenomena, including wake
interaction, drafting--kissing--tumbling dynamics, and clustering effects
\parencite{ladd_numerical_nodate,uhlmann_sedimentation_2014}.

In geophysical contexts, such interactions have been shown to strongly affect
crystal settling rates and spatial distributions in magma chambers
\parencite{martin_crystal_1988,weinstein_evolution_1988,martin_fluid-dynamical_1989}.
Accurately capturing these effects therefore requires resolving the full flow
field around multiple interacting crystals rather than relying on isolated
particle approximations.

From a surrogate modeling perspective, the Stokes regime has two useful properties. First, the absence of inertia leads to smooth, steady-state flow
fields for fixed crystal configurations. Second, incompressibility imposes a
strong global constraint on admissible velocity fields, which can be exploited
through appropriate output representations.

% ==========================================================
\section{Stream-Function Formulation in Two Dimensions}
\label{sec:stream_function}
% ==========================================================

In two-dimensional incompressible flow, the divergence-free condition
$\nabla \cdot \mathbf{u} = 0$ can be satisfied identically by introducing a
scalar stream function $\psi(x,z)$ such that
\begin{equation}
u_x = \frac{\partial \psi}{\partial z},
\qquad
u_z = -\frac{\partial \psi}{\partial x}.
\end{equation}
This transformation reduces the vector-valued velocity field to a single
scalar potential and guarantees incompressibility by construction.
% TODO: Klassische Fluid-Mechanik-Referenz ergänzen, z.\,B. Batchelor (1967) ``An Introduction to Fluid Dynamics'' oder Kim \& Karrila (1991) ``Microhydrodynamics'' -- noch nicht in references.bib

The stream-function formulation has a long tradition in classical
hydrodynamics, particularly for planar Stokes flow problems involving rigid
inclusions and sedimentation \parencite{ladd_numerical_nodate,uhlmann_sedimentation_2014}.
% TODO: Lehrbuch-Referenz ergänzen, z.\,B. Kim \& Karrila (1991) ``Microhydrodynamics'' -- noch nicht in references.bib
Streamlines
coincide with level sets of $\psi$, allowing qualitative flow features such as
recirculation zones, wakes, and stagnation points to be identified directly.

From a numerical and learning perspective, the stream function offers several
advantages. As a scalar field, $\psi$ typically exhibits smoother spatial
structure than individual velocity components, as it integrates information
over spatial derivatives. Moreover, representing the solution in terms of
$\psi$ eliminates the need to explicitly enforce the incompressibility
constraint during learning, in contrast to direct velocity-based approaches.

These properties motivate the use of the stream function as a learning target
in this thesis, particularly in comparison to direct prediction of velocity
components.

% ==========================================================
\section{Vorticity and Poisson Reconstruction of the Stream Function}
\label{sec:poisson_psi}
% ==========================================================

In two-dimensional incompressible flow, the scalar vorticity is defined as the
$z$-component of the curl of the velocity field,
\begin{equation}
\omega = \frac{\partial u_z}{\partial x}
       - \frac{\partial u_x}{\partial z},
\end{equation}
and measures the local rate of rotation of a fluid element. To derive the
governing equation for $\omega$, one applies the curl operator to the Stokes
momentum equation (neglecting the constant body force $\rho\mathbf{g}$, whose
curl vanishes):
\begin{equation}
\nabla \times \bigl(-\nabla p + \mu\,\nabla^2 \mathbf{u}\bigr) = \mathbf{0}.
\end{equation}
Since the curl of any gradient is identically zero, $\nabla \times \nabla p =
\mathbf{0}$, the pressure term drops out. Using the commutativity of curl and
the Laplacian for smooth fields, the remaining term gives
\begin{equation}
\mu\,\nabla^2 \omega = 0,
\end{equation}
showing that the vorticity satisfies a homogeneous Laplace equation in Stokes
flow.

A direct relation between $\omega$ and the stream function $\psi$ follows from
substituting the definitions $u_x = \partial\psi/\partial z$ and
$u_z = -\partial\psi/\partial x$ into the expression for vorticity:
\begin{equation}
\omega
= \frac{\partial u_z}{\partial x} - \frac{\partial u_x}{\partial z}
= -\frac{\partial^2 \psi}{\partial x^2} - \frac{\partial^2 \psi}{\partial z^2}
= -\nabla^2 \psi.
\end{equation}
Rearranging yields the Poisson equation for the stream function,
\begin{equation}
\label{eq:poisson_psi}
\nabla^2 \psi = -\omega.
\end{equation}

This relationship provides a natural link between velocity-based numerical
solutions and the stream-function representation. Given a divergence-free
velocity field from a numerical solver, the vorticity can be computed by
finite differences, and the stream function can be reconstructed by solving
Equation~\eqref{eq:poisson_psi} with appropriate boundary conditions.
In the present work, this procedure is applied to high-fidelity LaMEM
simulations to construct $\psi$ as a learning target.

It is important to emphasize that this Poisson solve does not introduce new
physics. Instead, it constitutes an auxiliary transformation that maps the
numerical reference solution into a representation that is more amenable to
learning. Similar transformations are commonly used in classical fluid
mechanics to analyze Stokes flow around particles and inclusions \parencite{ladd_numerical_nodate}.
% TODO: Ggf. Lehrbuch-Referenz ergänzen (Pozrikidis 1992 ``A Practical Guide to Boundary Element Methods'' o.\,ä.) -- noch nicht in references.bib


% ==========================================================
\section{Learning Targets as Physically Motivated Inductive Biases}
\label{sec:inductive_bias}
% ==========================================================

From a machine-learning perspective, the choice of output representation
introduces an implicit \emph{inductive bias}, i.e.\ a structural constraint on
the space of functions that the model can represent.
In the context of surrogate modeling for incompressible flow, this choice
strongly influences the degree to which physical structure must be learned
from data rather than enforced analytically.

Direct velocity prediction treats the learning task as an unconstrained
regression problem on a vector field. Incompressibility and momentum balance
are not enforced explicitly but must be inferred from the training data. While
this approach works well in many CNN-based flow surrogates
\parencite{ribeiro_deepcfd_nodate,chen_u-net_nodate,thuerey_deep_2020}, it can lead to
small but systematic violations of physical constraints.

Predicting the stream function $\psi$ introduces a stronger inductive bias by
restricting the hypothesis space to divergence-free velocity fields by
construction. This approach is conceptually related to divergence-free neural
representations proposed in the context of physics-informed and
structure-preserving learning \parencite{richter-powell_neural_2022}, and is
consistent with the neural operator framework \parencite{kovachki_neural_nodate}
that views surrogate models as approximations of solution operators on
function spaces. By reducing
the effective degrees of freedom of the output, the learning problem is
simplified and numerical conditioning can be improved.

Residual learning further modifies the inductive bias by decomposing the
solution into a physically motivated baseline and a learned correction. If the
baseline captures the dominant large-scale behavior of the flow, the residual
field is typically smoother and of lower amplitude. Such strategies have been
shown to improve learning efficiency and generalization in physics-based
surrogate models \parencite{sun_surrogate_2020}.

% ==========================================================
\section{U-Net Architecture}
\label{sec:unet}
% ==========================================================

The U-Net \parencite{ronneberger_u-net_2015} is an encoder--decoder convolutional neural
network originally developed for biomedical image segmentation. Its
architecture consists of three principal components: a contracting encoder
path, an expanding decoder path, and skip connections that bridge
corresponding resolution levels. Figure~\ref{fig:unet_arch} provides a
schematic overview of the architecture as used in this work.

\begin{figure}[htbp]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
  >=stealth,
  enc/.style  = {draw, fill=blue!12, rounded corners=3pt,
                 minimum width=1.4cm, minimum height=2.6cm,
                 align=center, font=\normalsize},
  bot/.style  = {draw, fill=red!12, rounded corners=3pt,
                 minimum width=1.4cm, minimum height=2.6cm,
                 align=center, font=\normalsize},
  dec/.style  = {draw, fill=teal!12, rounded corners=3pt,
                 minimum width=1.4cm, minimum height=2.6cm,
                 align=center, font=\normalsize},
  io/.style   = {align=center, font=\normalsize},
  skip/.style = {->, dashed, thick, gray!70},
  flow/.style = {->, thick},
]

% dx = horizontal step per level, dy = vertical step per level
% Block height = 2.6cm, dy = 3.2cm  =>  0.6cm gap between adjacent levels
\def\dx{2.6}
\def\dy{3.2}

% ---- Encoder (staircase: down-right) ----
\node[enc] (e1) at (0*\dx,  4*\dy) {Conv\\Block\\[4pt]32\,ch};
\node[enc] (e2) at (1*\dx,  3*\dy) {Down\\Block\\[4pt]64\,ch};
\node[enc] (e3) at (2*\dx,  2*\dy) {Down\\Block\\[4pt]128\,ch};
\node[enc] (e4) at (3*\dx,  1*\dy) {Down\\Block\\[4pt]256\,ch};

% ---- Bottleneck ----
\node[bot] (bot) at (4*\dx, 0)     {Down\\Block\\[4pt]512\,ch};

% ---- Decoder (staircase: up-right) ----
\node[dec] (d4) at (5*\dx,  1*\dy) {Conv\\Block\\[4pt]256\,ch};
\node[dec] (d3) at (6*\dx,  2*\dy) {Conv\\Block\\[4pt]128\,ch};
\node[dec] (d2) at (7*\dx,  3*\dy) {Conv\\Block\\[4pt]64\,ch};
\node[dec] (d1) at (8*\dx,  4*\dy) {Conv\\Block\\[4pt]32\,ch};

% ---- Input / Output ----
\node[io] (inp) at (-2.4, 4*\dy)
    {Mask, SDF,\\ $x$-Coord, $z$-Coord\\ (4\,channels)};
\node[io] (out) at (8*\dx + 2.2, 4*\dy)
    {$\hat{\psi}$\\ (1\,channel)};

% ---- Input → enc1 ----
\draw[flow] (inp.east) -- (e1.west);

% ---- Encoder diagonal arrows (south-east → north-west of next block) ----
\draw[flow] (e1.south east) --
    node[right, font=\small, xshift=2pt] {stride\,2} (e2.north west);
\draw[flow] (e2.south east) --
    node[right, font=\small, xshift=2pt] {stride\,2} (e3.north west);
\draw[flow] (e3.south east) --
    node[right, font=\small, xshift=2pt] {stride\,2} (e4.north west);
\draw[flow] (e4.south east) --
    node[right, font=\small, xshift=2pt] {stride\,2} (bot.north west);

% ---- Bottleneck → first decoder stage ----
\draw[flow] (bot.north east) --
    node[right, font=\small, xshift=2pt] {bilinear\,$\times\!2$} (d4.south west);

% ---- Decoder diagonal arrows ----
\draw[flow] (d4.north east) --
    node[right, font=\small, xshift=2pt] {bilinear\,$\times\!2$} (d3.south west);
\draw[flow] (d3.north east) --
    node[right, font=\small, xshift=2pt] {bilinear\,$\times\!2$} (d2.south west);
\draw[flow] (d2.north east) --
    node[right, font=\small, xshift=2pt] {bilinear\,$\times\!2$} (d1.south west);

% ---- dec1 → output ----
\draw[flow] (d1.east) --
    node[above, font=\small] {Conv\,$1{\times}1$} (out.west);

% ---- Skip connections (horizontal at each level, clear of all blocks) ----
\draw[skip] (e1.east) -- node[above, font=\small] {concat} (d1.west);
\draw[skip] (e2.east) -- node[above, font=\small] {concat} (d2.west);
\draw[skip] (e3.east) -- node[above, font=\small] {concat} (d3.west);
\draw[skip] (e4.east) -- node[above, font=\small] {concat} (d4.west);

\end{tikzpicture}%
}
\caption{Architecture of the U-Net surrogate model used in this work. The
encoder path (blue) descends the staircase from left to right, halving the
spatial resolution at each stage via strided convolutions while doubling the
channel depth (32 to 512). Skip connections (dashed) copy encoder feature maps
horizontally to the matching decoder level, where they are concatenated with
the upsampled signal. The decoder path (teal) ascends symmetrically via
bilinear upsampling. A final $1{\times}1$ convolution maps the 32-channel
output to the single-channel stream-function prediction $\hat{\psi}$.}
\label{fig:unet_arch}
\end{figure}

\subsection{Encoder Path}

The encoder successively reduces the spatial resolution of the input through
repeated application of convolutional layers followed by downsampling
operations (typically max-pooling or strided convolutions). At each stage, the
number of feature channels is increased, enabling the network to learn
increasingly abstract and spatially coarse representations. For an input of
spatial dimension $H \times W$, $L$ encoding stages produce feature maps at
resolutions $H/2^l \times W/2^l$ for $l = 1, \ldots, L$.

\subsection{Decoder Path and Skip Connections}

The decoder mirrors the encoder structure and progressively recovers spatial
resolution through upsampling operations (transposed convolutions or
interpolation) followed by convolutional layers. At each decoder stage, the
upsampled feature map is concatenated with the corresponding encoder feature
map via a skip connection. These skip connections are critical: they provide
the decoder with high-resolution spatial information that would otherwise be
lost during the encoding process, enabling the network to produce spatially
precise outputs.

\subsection{Multiscale Feature Extraction}

The hierarchical structure of the U-Net endows it with an inherently
multiscale receptive field. Shallow layers operate at fine spatial resolution
and capture localized features such as boundary layers and sharp gradients
near crystal surfaces. Deep layers operate at coarse resolution with large
effective receptive fields and can represent domain-scale flow structures.
The skip connections allow the final prediction to integrate information
across all scales simultaneously.

For multi-crystal Stokes flow, this multiscale property is directly relevant:
localized velocity gradients near crystal boundaries coexist with long-range
hydrodynamic interactions that span the entire computational domain. The
U-Net's ability to process both scales within a single forward pass makes it a
natural candidate for this class of problems.

% ==========================================================
\section{Fourier Neural Operator}
\label{sec:fno}
% ==========================================================

The Fourier Neural Operator (FNO) \parencite{li_fourier_2021} belongs to the family
of neural operators that learn mappings between infinite-dimensional function
spaces. Unlike conventional convolutional networks, which learn local spatial
kernels, the FNO parameterizes integral kernel operators in the spectral
domain.

\subsection{Fourier Layer}

The core building block of the FNO is the Fourier layer, which applies a
linear transformation in frequency space. Given an input function
$v(x) \in \mathbb{R}^{d_v}$ defined on a spatial domain, the Fourier layer
computes
\begin{equation}
\label{eq:fourier_layer}
v_{l+1}(x) = \sigma\!\Bigl(W_l\, v_l(x) + \mathcal{F}^{-1}\!\bigl(R_l \cdot \mathcal{F}(v_l)\bigr)(x)\Bigr),
\end{equation}
where $\mathcal{F}$ and $\mathcal{F}^{-1}$ denote the Fast Fourier Transform
and its inverse, $R_l \in \mathbb{C}^{k_{\max} \times d_v \times d_v}$ is a
learnable weight tensor applied to the lowest $k_{\max}$ Fourier modes, $W_l$
is a pointwise linear transformation (bias path), and $\sigma$ is a nonlinear
activation function.

The key operation is the multiplication $R_l \cdot \mathcal{F}(v_l)$ in
Fourier space, which is equivalent to a global convolution in physical space.
By retaining only $k_{\max}$ modes, the FNO implicitly applies a low-pass
filter while keeping the number of learnable parameters independent of the
spatial discretization. Figure~\ref{fig:fourier_layer} illustrates the
structure of a single Fourier layer.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  >=stealth,
  box/.style  = {draw, fill=white, rounded corners=2pt,
                 minimum width=2.0cm, minimum height=0.75cm,
                 align=center, font=\small},
  circ/.style = {draw, circle, minimum size=0.7cm, inner sep=0pt,
                 align=center, font=\normalsize},
  io/.style   = {align=center, font=\small},
  arr/.style  = {->, thick},
  seg/.style  = {thick},
]

\def\ytop{1.8}
\def\ybot{-1.8}

% Outer box (drawn first → behind all nodes)
\draw[rounded corners=5pt, fill=gray!5, draw=gray!50]
    (1.3, -2.4) rectangle (15.2, 2.4);
\node[font=\small\itshape, anchor=south west] at (1.5, -2.3)
    {Fourier Layer};

% Input
\node[io] (in) at (0, 0) {$v_l(x)$};

% Spectral path
\node[box] (fft)  at (3.5,  \ytop) {$\mathcal{F}$};
\node[box] (rk)   at (6.8,  \ytop)
    {$R_l\,\cdot$\\[-3pt]\scriptsize($k_{\max}$ modes)};
\node[box] (ifft) at (10.0, \ytop) {$\mathcal{F}^{-1}$};

% Pointwise path
\node[box] (wl)   at (6.8,  \ybot) {$W_l$};

% Sum and activation
\node[circ] (sum) at (12.2, 0) {$+$};
\node[box]  (sig) at (14.0, 0) {$\sigma$};

% Output
\node[io] (out) at (15.8, 0) {$v_{l+1}(x)$};

% Path labels
\node[font=\scriptsize, gray, anchor=south west] at (2.2, 0.15)
    {spectral};
\node[font=\scriptsize, gray, anchor=north west] at (2.2, -0.15)
    {pointwise};

% Branch point
\coordinate (split) at (2.0, 0);
\fill (split) circle (1.8pt);

% Connections
\draw[seg] (in.east) -- (split);
\draw[arr] (split)   |- (fft.west);    % up → spectral path
\draw[arr] (split)   |- (wl.west);     % down → pointwise path
\draw[arr] (fft)     --  (rk);
\draw[arr] (rk)      --  (ifft);
\draw[arr] (ifft.east) -| (sum.north); % spectral → sum (right then down)
\draw[arr] (wl.east)   -| (sum.south); % pointwise → sum (right then up)
\draw[arr] (sum) -- (sig);
\draw[arr] (sig) -- (out);

\end{tikzpicture}
\caption{Structure of a single Fourier layer
(Equation~\eqref{eq:fourier_layer}). The input $v_l(x)$ is split into two
parallel paths. The spectral path (top) applies the Fast Fourier Transform
$\mathcal{F}$, multiplies the lowest $k_{\max}$ Fourier modes by the
learnable weight tensor $R_l$, and transforms back via $\mathcal{F}^{-1}$,
performing a global convolution in physical space. The pointwise path
(bottom) applies the local linear transformation $W_l$ directly in the
spatial domain. Both outputs are summed and passed through a nonlinear
activation $\sigma$.}
\label{fig:fourier_layer}
\end{figure}

\subsection{Architecture}

A complete FNO model consists of three stages:
\begin{enumerate}
    \item A pointwise lifting layer $P: \mathbb{R}^{d_a} \to \mathbb{R}^{d_v}$ that projects the input channels into a higher-dimensional feature space.
    \item A sequence of $L$ Fourier layers as defined in Equation~\eqref{eq:fourier_layer}, each combining spectral and local processing.
    \item A pointwise projection layer $Q: \mathbb{R}^{d_v} \to \mathbb{R}^{d_o}$ that maps the learned features to the desired output dimension.
\end{enumerate}

\subsection{Properties Relevant to Stokes Flow}

The FNO architecture has several properties that are relevant for the
surrogate modeling task considered in this work:

\begin{itemize}
    \item \textbf{Global receptive field.} Each Fourier layer captures
    dependencies across the entire spatial domain in a single operation. For
    Stokes flow, where the elliptic character of the governing equations
    implies that local perturbations propagate globally, this is a structurally
    natural property.
    \item \textbf{Spectral bias.} The truncation to $k_{\max}$ Fourier modes
    introduces an implicit smoothness prior. While this favors the
    representation of large-scale flow structures, it may limit accuracy near
    sharp interfaces such as crystal boundaries, where high-frequency content
    is physically significant \parencite{qin_toward_2024}.
    \item \textbf{Discretization invariance.} Because the FNO operates on
    continuous function representations via their Fourier coefficients, a model
    trained on one grid resolution can, in principle, be evaluated on a
    different resolution without retraining.
\end{itemize}

% ==========================================================
\section{Architectural Comparison: U-Net vs.\ FNO}
\label{sec:arch_comparison}
% ==========================================================

The U-Net and FNO embody fundamentally different inductive biases for
structured prediction tasks. Table~\ref{tab:arch_comparison} summarizes the
key distinctions.

\begin{table}[htbp]
\centering
\caption{Structural comparison of U-Net and FNO architectures.}
\label{tab:arch_comparison}
\begin{tabular}{lll}
\toprule
\textbf{Property} & \textbf{U-Net} & \textbf{FNO} \\
\midrule
Domain of operation & Spatial & Spectral (Fourier) \\
Receptive field & Hierarchical, local-to-global & Global per layer \\
Multiscale mechanism & Encoder--decoder + skip connections & Mode truncation \\
High-frequency content & Preserved via skip connections & Limited by $k_{\max}$ \\
Parameter scaling & Depends on depth and channels & Independent of resolution \\
\bottomrule
\end{tabular}
\end{table}

For multi-crystal Stokes flow, these differences translate into complementary
strengths: the U-Net is expected to excel at resolving sharp boundary layers
and localized wake structures due to its hierarchical spatial processing,
while the FNO may offer advantages in capturing the global, elliptic character
of Stokes interactions through its spectral representation. The empirical
comparison of these architectures on identical training data and evaluation
metrics is a central contribution of this thesis.

% ==========================================================
\section{Implications for Multi-Crystal Sedimentation Modeling}
\label{sec:theory_implications}
% ==========================================================

The theoretical considerations discussed above have direct implications for
surrogate modeling of multi-crystal sedimentation. Long-range hydrodynamic
interactions and incompressibility impose global structure on the flow field,
which is difficult to capture through purely local regression of velocity
components. Predicting the stream function $\psi$ provides a principled way to
reduce the complexity of the learning task by enforcing incompressibility
analytically.

The choice of network architecture interacts with this physical structure in
complementary ways. The U-Net's hierarchical encoder--decoder processing
naturally mirrors the multiscale character of the flow: sharp boundary layers
near crystal surfaces are captured at fine resolution levels, while
domain-scale circulation patterns are represented in the deep, coarse layers.
The FNO's spectral approach, on the other hand, directly reflects the
elliptic nature of the Stokes equations, where local perturbations propagate
globally—a property that is inherently encoded in the Fourier representation.

These complementary architectural biases are what motivates the comparison
pursued in this thesis: by training both architectures on the same data and
evaluating them under identical conditions, we can disentangle the effects of
physical output representation (stream function) from those of architectural
design (spatial vs.\ spectral processing).

The next chapter builds on this theoretical foundation by translating these
considerations into a concrete computational methodology for data generation,
network design, training, and evaluation.
