\chapter{Methodology}
\label{ch:methodology}

This chapter describes the computational workflow for generating training data, constructing machine-learning inputs, training surrogate models, and evaluating their generalization performance. The methodology builds on the literature reviewed in Chapter~\ref{ch:related} and provides a reproducible end-to-end pipeline for structured-grid flow-field prediction.

Both surrogate architectures considered in this thesis—U-Net and Fourier Neural Operator (FNO)—predict the stream function $\psi$ on a fixed $256 \times 256$ grid. The velocity field is then recovered analytically from
$\psi$ via its spatial derivatives, guaranteeing incompressibility by construction. The simulation setup, input encoding, optimization procedure, and evaluation protocol are kept identical for both architectures, so that
performance differences can be attributed to the architectural design—spatial hierarchical processing (U-Net) versus spectral global processing (FNO)—rather than to confounding factors in data or training.

% ==========================================================
\section{Learning Target: Stream-Function Prediction}
\label{sec:learning_target}
% ==========================================================

Both surrogate architectures predict the scalar stream function $\psi(x,z)$ on the computational grid. The velocity field is recovered analytically via
\[
    \mathbf{u}=\left(\frac{\partial \psi}{\partial z},\;
    -\frac{\partial \psi}{\partial x}\right),
\]
which enforces incompressibility by construction. The theoretical motivation for this formulation and its advantages over direct velocity prediction were discussed in Chapter~\ref{ch:theory}.

From the perspective of surrogate modeling, predicting $\psi$ instead of $\mathbf{u}$ directly offers two key benefits. First, it reduces the learning problem from a vector-valued to a scalar regression task, lowering the
effective degrees of freedom. Second, incompressibility is guaranteed analytically rather than having to be learned from data, which eliminates a common source of systematic error in purely data-driven velocity surrogates
\parencite{ribeiro_deepcfd_nodate,thuerey_deep_2020}. This choice injects physical structure into the learning problem through the output representation rather than through penalty terms, consistent with divergence-free parameterizations advocated in related work \parencite{richter-powell_neural_2022}.

The central experimental factor in this thesis is therefore not the choice of learning target—which is fixed to $\psi$ for all experiments—but the choice of network architecture. By training both U-Net and FNO on identical data with the same learning target, performance differences can be attributed to the architectural inductive bias: spatial hierarchical processing versus spectral global processing.

% ==========================================================
\section{LaMEM Simulations for Data Generation}
\label{sec:lamem_generation}
% ==========================================================

Training data are generated using the geodynamic finite-difference code \textsc{LaMEM}, which solves the incompressible Stokes equations on a regular Cartesian grid. The simulation setup follows established geodynamic Stokes
discretization practices, where robustness under viscosity and density contrasts is essential \parencite{thieulot_choice_2022}.  

Each simulation represents a steady (quasi-static) configuration of $1\ldots N_{\max}$ rigid crystals settling in a viscous matrix. Since the incompressible Stokes equations are quasi-static, each simulation directly yields a steady-state velocity field for a fixed crystal arrangement. Transient crystal dynamics are not modeled; the surrogate is trained on individual steady-state flow snapshots.

\subsection{Domain and Discretization}

The computational domain is a rectangular box discretized on a fixed $256\times256$ grid with uniform spacing $\Delta x=\Delta z$. The surrogate models operate on fields represented on this grid. While physical units are
retained internally during simulation, learning is performed on consistently scaled grid-based fields, and the fixed discretization provides a controlled benchmark for comparing output representations.

Crystals are modeled as rigid circular inclusions with prescribed radii and randomized positions. The number of crystals $N$ is drawn uniformly from $\{1,\ldots,N_{\max}\}$ to expose the surrogate to a broad range of interaction regimes, from isolated single-crystal wakes to strongly interacting clusters. Such multi-particle interaction effects are well documented in classical sedimentation studies \parencite{ladd_numerical_nodate} and
motivate an explicit evaluation of generalization across particle number. Randomization is performed independently for each simulation run, such that the training distribution samples a diverse subset of the configuration space.

\subsection{Material Properties and Governing Equations}

The matrix phase is assigned viscosity $\mu$ and density $\rho_f$. Crystals are approximated as effectively rigid particles by assigning a viscosity several orders of magnitude larger than $\mu$ and a density $\rho_p>\rho_f$ to induce gravitational settling. Within LaMEM, this results in a two-phase Stokes problem with sharp viscosity and density contrasts.

The governing equations are the incompressible Stokes equations
\[
\mathbf{0}=-\nabla p + \mu\nabla^2\mathbf{u} + \rho\mathbf{g},
\qquad
\nabla\cdot\mathbf{u}=0,
\]
solved with free-slip boundary conditions on all walls. Free-slip is enforced by vanishing normal velocity and vanishing tangential shear stress at the boundaries, reducing artificial boundary-layer effects and providing a simple, well-posed benchmark setting.

\subsection{Extraction of Flow Quantities}

For each simulation, LaMEM outputs grid-based fields including the phase indicator and the velocity components as well as velocity gradients:
\[
\{\text{phase field},\,u_x,\,u_z,\,\nabla u_x,\,\nabla u_z\}.
\]
The phase field indicates whether a cell belongs to the crystal or matrix.

From the velocity gradients, the scalar vorticity is computed as
\[
\omega=\frac{\partial u_z}{\partial x}-\frac{\partial u_x}{\partial z},
\]
consistent with the two-dimensional convention used in Chapter~\ref{ch:theory}. The vorticity field serves as the right-hand side for reconstructing the stream function $\psi$ via a Poisson solve, as described in Section~\ref{sec:psi}.

% ==========================================================
\section{Computing and Normalizing the Stream Function}
\label{sec:psi}
% ==========================================================

The stream function is reconstructed from vorticity by solving the Poisson equation
\[
\Delta\psi=-\omega
\]
on the same $256\times256$ grid. Reconstructing $\psi$ via a Poisson problem is standard in two-dimensional incompressible low-Reynolds-number flow theory \parencite{happel_low_1983}. In the present workflow, this Poisson solve is used only to define a physically consistent scalar learning target; the underlying reference solution remains the LaMEM Stokes solution.

\subsection{Discretization and Boundary Conditions}

The Laplace operator is discretized with a second-order five-point stencil on interior grid points. Denoting the one-dimensional second-difference matrices in $x$ and $z$ as $T_x \in \mathbb{R}^{n_x \times n_x}$ and $T_z \in \mathbb{R}^{n_z \times n_z}$, respectively, the discrete Laplacian is assembled via Kronecker products,
\[
    L = \frac{1}{\Delta x^2}\,(I_z \otimes T_x)
      + \frac{1}{\Delta z^2}\,(T_z \otimes I_x),
\]
where $I_x$ and $I_z$ are identity matrices of matching dimension. This construction yields a sparse $(n_x n_z)\times(n_x n_z)$ system and is standard for uniform-grid Poisson problems.

Homogeneous Dirichlet conditions $\psi|_{\partial\Omega}=0$ are imposed on all boundary nodes by replacing the corresponding rows in $L$ with unit rows and setting the right-hand side entries to zero. This fixes the stream-function gauge and yields a unique discrete solution.

The resulting sparse linear system $L\,\psi = -\omega$ is solved directly using a sparse LU factorization (Julia's built-in backslash operator, which calls UMFPACK). Since the grid is fixed at $256\times256$, the system has $65\,536$ unknowns and the direct solve completes in well under one second per sample, making an iterative approach unnecessary. After solving, a constant offset equal to the value at the corner node $\psi_{1,1}$ is subtracted to remove any residual floating-point gauge ambiguity.

Since all evaluation comparisons are ultimately performed against LaMEM-derived quantities on the same grid, the Poisson solve serves as an auxiliary reconstruction step rather than an independent physical model.

\subsection{Motivation and Procedure for Normalization}

Raw values of $\psi$ can span several orders of magnitude (typically $10^{-16}$ to $10^{-10}$), depending on velocity magnitude, viscosity, and grid spacing. While such values are well resolved in the numerical solver, they are unfavorable for gradient-based learning. The theoretical reason is rooted in the geometry of the loss landscape: for a gradient-based learner, the condition number $\kappa = \lambda_{\max}/\lambda_{\min}$ of the Hessian of the loss with respect to the network weights determines how well a single learning rate can serve all directions of the weight space simultaneously. For a linear layer with input $x$, the Hessian equals the input covariance matrix $H = P^{-1}\sum_p x^p x^{p\top}$; consequently, inputs of very large magnitude generate a dominant eigenvalue $\lambda_{\max}$, inflating $\kappa$ and forcing the optimizer into inefficient zigzag steps along the ill-conditioned directions \parencite{LeCun2012}. In practice, this manifests as slow convergence and sensitivity to the choice of learning rate. Shifting inputs to zero mean and unit variance reduces $\kappa$ and yields a more isotropic error surface on which standard first-order optimizers such as Adam converge reliably.

To bring $\psi$ into a numerically suitable range without discarding its spatial structure or sign, each sample is normalized individually. We compute a characteristic order of magnitude
\[
p_{\mathrm{mean}}=
\mathrm{round}\!\left(
\mathrm{mean}\left(\log_{10}(|\psi|+\epsilon)\right)
\right),
\qquad \epsilon=10^{-30},
\]
and scale the field as
\[
\psi_{\mathrm{norm}}=\psi\cdot 10^{-p_{\mathrm{mean}}}.
\]
Training is performed on $\psi_{\mathrm{norm}}$, and predictions are rescaled after inference by multiplying with $10^{p_{\mathrm{mean}}}$. This preserves the spatial structure and sign of $\psi$ while shifting values into an $\mathcal{O}(1)$ range that yields stable optimization.

% ==========================================================
\section{Input Representation: Mask, Distance Field, and Coordinates}
\label{sec:input_channels}
% ==========================================================

Both architectures receive the same five-channel input, generated by a shared data pipeline. Training data are stored as JLD2 files and loaded identically for both U-Net and FNO training. Encoding geometry through masks, distance fields, and coordinate channels is common in structured-grid PDE surrogates and geometry-aware learning frameworks \parencite{thuerey_deep_2020,oldenburg_geometry_2022}.

The five input channels are:

\begin{enumerate}
    \item \textbf{Crystal mask.}
    A binary indicator field $M(x,z)\in\{0,1\}$ specifying crystal locations:
    \[
        M(x,z)=
        \begin{cases}
            1, & \text{inside any crystal},\\
            0, & \text{in the matrix}.
        \end{cases}
    \]
    This channel encodes the location of rigid inclusions and material contrasts.

    \item \textbf{Signed distance field (SDF).}
    For each grid point, the signed distance to the nearest crystal boundary is computed: negative inside crystals, positive outside. This provides a smooth measure of boundary proximity that complements the sharp binary mask and helps the network resolve steep velocity gradients near solid--fluid interfaces.

    \item \textbf{Radius of nearest crystal.}
    For each grid point, the radius of the geometrically closest crystal is stored as an explicit channel. In Experiments~1--3, all crystals share the same fixed radius ($r = 0.05$), so this channel carries no configurational variation and has negligible effect on model behavior. It is included as a forward-compatible feature for Experiment~4, in which crystal radii vary across the training and evaluation sets: by explicitly encoding the crystal size at each grid point, the network receives the information necessary to condition its prediction on the local geometry. Without this channel, a model trained on multiple discrete radii would be unable to infer the radius of the crystal in a given sample at inference time, limiting its ability to generalize to unseen sizes.

    \item \textbf{Normalized $x$-coordinate.}
    The horizontal coordinate is linearly mapped to $x_{\mathrm{norm}}\in[-1,1]$, providing the network with absolute positional information.

    \item \textbf{Normalized $z$-coordinate.}
    Analogously, the vertical coordinate is mapped to $z_{\mathrm{norm}}\in[-1,1]$. Together, the coordinate channels reduce symmetry-induced and position-dependent artifacts.
\end{enumerate}

Each sample is represented by an input tensor of shape $(256,256,5)$. The input encodes inclusion geometry, proximity to crystal boundaries, and domain position, and is identical for both architectures.

% ==========================================================
\section{Surrogate Architectures}
\label{sec:architectures}
% ==========================================================

Two architectures are compared in this thesis. Both receive the same five-channel input of shape $(256,256,5)$ and produce a single-channel output $\psi_{\mathrm{norm}}$ of shape $(256,256,1)$. The theoretical foundations
of both architectures were introduced in Chapter~\ref{ch:theory}; this section describes the specific configurations used in the experiments.

\subsection{U-Net}
\label{sec:unet_architecture}

The U-Net follows the encoder--decoder principle of the original architecture \parencite{ronneberger_u-net_2015} and common adaptations for laminar flow-field prediction on Cartesian grids \parencite{thuerey_deep_2020,chen_u-net_nodate}.

\paragraph{Encoder.}
The encoder consists of four resolution levels. Spatial resolution is reduced by a factor of two at each level, while feature widths increase as $32 \rightarrow 64 \rightarrow 128 \rightarrow 256$. Each level contains two
$3\times3$ convolutions with batch normalization and ReLU activations, followed by downsampling via a strided convolution (stride~2). Strided downsampling is used instead of max pooling to keep the operation learnable and to reduce grid-aligned artifacts in smooth Stokes-flow targets.

\paragraph{Bottleneck.}
At the coarsest resolution, a bottleneck block with two $3\times3$ convolutions encodes global flow structure conditioned on the entire crystal configuration. This compressed representation provides the long-range context required to model hydrodynamic interactions between distant inclusions.

\paragraph{Decoder.}
The decoder mirrors the encoder: spatial resolution is restored by bilinear interpolation, followed by a two-layer $3\times3$ convolutional block with batch normalization and ReLU activation. Bilinear interpolation is used instead of transposed convolutions to avoid checkerboard artifacts that arise from the periodic zero-insertion pattern of deconvolution \parencite{odena_deconvolution_2016}. Encoder feature maps are concatenated with decoder feature maps at matching resolutions through skip connections, preserving high-resolution geometric detail essential for reconstructing sharp structures near crystal boundaries while maintaining global consistency.

\paragraph{Output layer.}
A final $1\times1$ convolution projects the decoder features onto a single output channel ($\psi_{\mathrm{norm}}$). No activation is applied in the final layer, as the task is a continuous regression problem.

\subsection{Fourier Neural Operator}
\label{sec:fno_architecture}

The FNO follows the architecture proposed by \textcite{li_fourier_2021}, consisting of a pointwise lifting layer, a sequence of Fourier layers, and a pointwise projection layer.

\paragraph{Lifting.}
A pointwise linear layer maps the five input channels into a higher-dimensional feature space of width $d_v$.

\paragraph{Fourier layers.}
Each Fourier layer applies the operation defined in Equation~\eqref{eq:fourier_layer}: the input is transformed into Fourier space via a two-dimensional FFT, multiplied with a learnable complex weight tensor $R_l$ for the lowest $k_{\max}$ modes, and transformed back. This spectral path is combined with a pointwise linear transformation (bias path) and a nonlinear activation. A stack of $L$ such layers forms the core of the model.

\paragraph{Projection.}
A pointwise linear layer maps the learned feature representation to the single output channel $\psi_{\mathrm{norm}}$.

\paragraph{Hyperparameters.}
The specific values of $d_v$, $k_{\max}$, and $L$ are reported together with the experimental results in Chapter~\ref{ch:results}, where they are chosen to yield a model with comparable parameter count to the U-Net to ensure a fair comparison.

\subsection{Regularization and Practical Considerations}

Both architectures are deliberately kept lightweight to enable repeated training runs on standard GPU hardware. For the U-Net, batch normalization stabilizes optimization. For the FNO, the implicit spectral smoothness prior
imposed by mode truncation serves as an architectural regularizer. Generalization is primarily promoted through geometric diversity in the pre-generated training data and the physically structured output representation. Dropout is not used, as the large variety of randomized crystal configurations in the training set provides sufficient regularization.

% ==========================================================
\section{Training Procedure}
\label{sec:training}
% ==========================================================

\subsection{Data Splits and Benchmark Design}

The dataset is divided into three fixed splits—training, validation, and test—each generated once from randomized crystal configurations and stored on disk prior to any model training. All three splits are identical for both
the U-Net and FNO pipelines, ensuring that performance differences between architectures are not confounded by differences in training or evaluation data. The validation split is used for monitoring during training, while the
test split provides a reproducible benchmark for the final evaluation. 

\subsection{Batching and Device Handling}

Samples are stored in \texttt{.jld2} format to enable efficient loading of multi-channel arrays in Julia. Mini-batches are assembled on the fly, with batch size chosen as a compromise between gradient stability and GPU memory
constraints. Data tensors and model parameters are explicitly moved to CPU or GPU to ensure consistent device placement and portable execution.

\subsection{Loss Function}

The primary loss is the mean-squared error
\[
\mathcal{L}_{\mathrm{MSE}}=
\frac{1}{N_{\mathrm{pix}}}\sum_i
\left(y^{\mathrm{pred}}_i-y^{\mathrm{true}}_i\right)^2.
\]
Both architectures are trained with an extended composite loss that combines the MSE with gradient-matching and boundary terms, increasing robustness near sharp crystal boundaries and encouraging physically consistent velocity fields. The specific loss weights and warmup schedule are detailed in Chapter~\ref{ch:experiments}.

\subsection{Optimization and Checkpointing}

Model parameters are optimized using the Adam optimizer \parencite{kingma2017adammethodstochasticoptimization} with adaptive learning rates. Adam maintains exponential moving averages of both the gradient (first moment) and its element-wise square (second moment), and applies bias correction to both estimates. This yields per-parameter adaptive step sizes that are approximately bounded by the chosen learning rate $\alpha$, making optimization robust to varying gradient magnitudes across network layers and reducing sensitivity to the initial learning rate choice. A fixed learning rate is used within each run, without learning rate scheduling, to keep experiments comparable. After each epoch, model checkpoints are stored in a CPU-compatible format to enable reproducible evaluation on systems without GPU support. For each run, training configuration metadata (hyperparameters, data paths, and random seeds) are archived alongside the checkpoint.

% ==========================================================
\section{Evaluation Pipeline}
\label{sec:evaluation}
% ==========================================================

Model predictions are evaluated on the fixed test dataset containing crystal geometries not used during training. This separation ensures that evaluation captures generalization to configurations not seen during training.

Primary evaluation metrics include:
\begin{itemize}
    \item mean-squared error (MSE) of the predicted target field,
    \item relative $L_2$ errors for $\psi$, $\partial_x\psi$, and $\partial_z\psi$,
          which quantify errors in both amplitude and derived velocity gradients,
    \item pixelwise error thresholds $\epsilon_{0.01}$, $\epsilon_{0.05}$, and
          $\epsilon_{0.10}$, defined as the fraction of grid points exceeding
          relative error levels of $1\%$, $5\%$, and $10\%$.
\end{itemize}

Errors are grouped by the number of crystals in the test sample ($1\ldots N_{\max}$), enabling direct assessment of how prediction quality changes as hydrodynamic interactions become more complex—an approach motivated
by generalization studies that stratify performance across geometric complexity \parencite{morimoto_generalization_2022}. The use of MSE and relative norms follows common practice in benchmarking data-driven CFD surrogates
\parencite{takamoto_pdebench_nodate,lu_comprehensive_2022}. The same evaluation framework is applied identically to both U-Net and FNO predictions, ensuring that all reported differences reflect architectural properties rather than evaluation artifacts.

% ==========================================================
\section{Summary}
\label{sec:method_summary}
% ==========================================================

This chapter described the full surrogate modeling workflow: LaMEM-based data generation, reconstruction and normalization of the stream function, input encoding, U-Net and FNO architecture configurations, training procedure, and evaluation protocol. The modular design ensures that all pipeline components except the network architecture are kept identical for both models, so that performance differences can be attributed to the architectural inductive bias rather than to differences in data or training. The next chapter describes the technical implementation of this pipeline.
