% ----------------------------------------------------------------------------
% Chapter 10: Conclusion
% ----------------------------------------------------------------------------
%
% ============================================================
% QUELLENVORSCHLÄGE FÜR DIESES KAPITEL
% ============================================================
% Zusammenfassung der Beiträge (verweist auf Literatur, die du übertroffen/ergänzt hast):
%   \parencite{thuerey_deep_2020}            -- Bisheriger Stand: U-Net für variable Geometrien
%   \parencite{li_fourier_2021}              -- Bisheriger Stand: FNO für PDE-Surrogate
%   \parencite{lu_comprehensive_2022}        -- Fairer Architekturvergleich als Referenzpunkt
%   \parencite{morimoto_generalization_2022} -- Generalisierungsanalyse als methodischer Vorläufer
%
% Ausblick / Zukünftige Arbeiten:
%   \parencite{wen_u-fnoenhanced_2022}       -- U-FNO: Hybridarchitektur als nächster Schritt
%   \parencite{qin_toward_2024}              -- Verbesserung Spektralbias (adaptive Modes)
%   \parencite{sanchez-gonzalez_learning_nodate} -- GNN für größere Partikelzahlen
%   \parencite{zhu_bayesian_2018}            -- Unsicherheitsquantifizierung
%   \parencite{hu_applying_2023}             -- PINN-Erweiterung für physikkonsistente Vorhersagen
%   \parencite{koehler_apebench_nodate}      -- Autoregressive Erweiterungen (zeitabhängige Settl.)
%   \parencite{li_fourier_nodate}            -- Geo-FNO für irreguläre Kristallgeometrien
%   \parencite{patocka_settling_2020}        -- Breitere geodynamische Anwendbarkeit
% ============================================================
%
\chapter{Conclusion and Outlook}
\label{ch:conclusions}

% ============================================================================
\section{Summary of Findings}
% ============================================================================

This thesis investigated the ability of two neural network architectures—the U-Net and the Fourier Neural Operator—to serve as data-driven surrogate models for two-dimensional Stokes flow around settling crystals. The key question was whether these models can generalize across varying numbers and spatial arrangements of crystals, and which architecture is better suited for this task. The main findings are summarized below.

In the single-crystal baseline (Experiment~1), the FNO consistently outperforms the U-Net: the best FNO configurations achieve mean relative $L_2$ errors of $1.6$--$2.3\%$ on the stream function~$\psi$, compared to $5.5$--$6.5\%$ for the best U-Net configurations. The FNO also shows substantially more stable training dynamics, while the U-Net exhibits pronounced validation loss oscillations at higher learning rates, particularly with batch size~16. The adoption of a composite loss function for U-Net training effectively reduces the velocity error relative to the $\psi$ error for batch size~8 (from a factor of roughly $3\times$ to $1.5$--$2\times$), but leads to increased errors at batch size~16, highlighting the sensitivity of multi-term loss functions to the training configuration.

In the multi-crystal generalization experiments (Experiments~2 and~3), the FNO achieves in-distribution errors of approximately $4$--$9\%$ on~$\psi$, maintaining a roughly $3$--$4\times$ advantage over the U-Net ($17$--$29\%$). OOD generalization is strongly influenced by the training distribution: in Experiment~2 (trained on $n \leq 10$), errors exceed $10\%$ already at $n = 20$ (twice the training maximum), while in Experiment~3 (trained on $n \leq 25$) the FNO maintains errors broadly comparable to its in-distribution performance up to $n = 35$. This demonstrates that the training distribution width is a key factor determining the OOD generalization range, consistent with findings by \textcite{morimoto_generalization_2022} for multi-obstacle flow surrogates. The U-Net shows a more uniform error profile across crystal counts, but at a consistently high error level that reflects a general difficulty in capturing the superposition of multi-crystal hydrodynamic interactions.

The stream-function output representation proves to be an effective inductive bias: incompressibility is guaranteed analytically for all predictions, confirmed by divergence RMS values on the order of $10^{-26}$, and no mass-conservation penalty term is needed during training.

Experiment~4 reveals an important limitation of the FNO for out-of-distribution crystal sizes. When evaluated on crystals larger than any seen during training, the FNO degrades severely ($71.3 \pm 4.8\%$ on~$\psi$), while the U-Net shows substantially lower errors ($44.9 \pm 8.2\%$). This represents a reversal of the architectural ranking observed in Experiments~1--3, and can be attributed to the spectral truncation at $k_{\max} = 16$: large crystals produce boundary-layer structures that require higher effective spatial frequencies than the fixed Fourier modes can represent. For small out-of-distribution crystals, both architectures perform comparably ($\approx 11$--$12\%$ on~$\psi$), suggesting that the FNO's spectral limitation is specifically triggered by geometric upscaling.

% ============================================================================
\section{Addressing the Research Questions}
% ============================================================================

The central research question asked to what extent U-Net and FNO architectures can generalize to unseen sedimentation scenarios with varying particle numbers and spatial arrangements, and which factors fundamentally limit this generalization.

The results show that both architectures can generalize to some extent beyond their training distributions, but in qualitatively different ways. The FNO generalizes well up to a crystal count that is roughly $1.2$--$1.5\times$ the training maximum, after which errors increase sharply. The generalization range can be extended by training on a wider distribution of crystal counts, as demonstrated by the comparison between Experiments~2 and~3. The U-Net, by contrast, shows no abrupt OOD failure but struggles already within the training distribution due to its limited ability to capture long-range hydrodynamic interactions.

The three hypotheses stated in Chapter~\ref{ch:introduction} are partially confirmed. The first hypothesis—that sufficient sampling of the geometric configuration space enables generalization—is confirmed for the FNO but only weakly for the U-Net, which shows high errors even at training crystal counts. The second hypothesis—that predicting the stream function rather than the velocity components directly reduces task complexity by converting a vector regression problem into a scalar one and enforcing incompressibility analytically by construction—is confirmed by the divergence-free predictions and the improved velocity accuracy relative to plain velocity prediction. The third hypothesis—that multiscale or spectral architectures can capture crystal--fluid interactions—is supported by the FNO's superior in-distribution accuracy, while the U-Net's local hierarchical processing proves less effective for this globally elliptic problem. Experiment~4 adds an important qualification: the spectral architecture's advantage does not extend to geometric size generalization, where the fixed Fourier mode truncation becomes a bottleneck.

% ============================================================================
\section{Outlook}
% ============================================================================

Several directions for future work emerge from the results of this thesis.

\paragraph{Extended FNO training for Experiments 2 and 3.}
The FNO results for Experiments~2 and~3 are based on 100-epoch training runs. Re-training to 200~epochs would provide a fairer basis for the architectural comparison in the multi-crystal regime and potentially improve the reported FNO error levels further.

\paragraph{Adaptive Fourier modes for size generalization.}
Experiment~4 shows that the FNO's spectral truncation at $k_{\max} = 16$ is the primary factor limiting its performance for large out-of-distribution crystals. Architectures with adaptive or multi-resolution frequency representations—such as the Geo-FNO \parencite{li_fourier_nodate} or convolutional neural operators \parencite{raonic_convolutional_2023}—could potentially overcome this limitation and enable robust generalization across crystal sizes.

\paragraph{Addressing U-Net capacity limitations in dense multi-crystal settings.}
The high-frequency artifacts in U-Net predictions for dense crystal configurations reflect an architectural capacity constraint rather than a training issue. Potential remedies include increasing the base channel width or adding an additional encoder level to expand the model's representational capacity, incorporating attention mechanisms to provide explicit long-range context without relying solely on the pooling hierarchy, or replacing the fixed convolutional architecture with a hybrid that combines local U-Net processing with global spectral layers as in U-FNO \parencite{wen_u-fnoenhanced_2022}.

\paragraph{Hybrid architectures.}
The U-FNO \parencite{wen_u-fnoenhanced_2022} combines the FNO's global spectral processing with U-Net-style local branches, potentially addressing the FNO's known limitation near sharp interfaces while retaining its generalization properties. Applying this architecture to the multi-crystal Stokes flow problem would be a natural extension.

\paragraph{Post-processing spectral filtering for velocity recovery.}
A practical limitation of the stream-function approach is that finite-difference differentiation of~$\hat\psi$ amplifies high-frequency residuals in the predicted field into proportionally larger velocity errors. The gradient-matching loss term in the composite loss partially mitigates this issue, but at the cost of additional hyperparameter tuning. A complementary and architecturally independent approach is to apply a low-pass spectral filter to $\hat\psi$ before computing velocity derivatives. Since physically meaningful Stokes flow fields are smooth outside crystal boundaries, a filter that attenuates wavenumbers above a threshold $k_\text{cut}$ would suppress numerical noise without distorting the flow structure. The optimal threshold $k_\text{cut}$ and filter shape could be determined from the known spectral content of the training data, potentially as a learned post-processing step.

\paragraph{Three-dimensional and time-dependent settings.}
Extending the surrogate modeling approach to three-dimensional Stokes flow or to time-dependent settling dynamics would be a major step towards geophysically realistic applications. Graph neural network approaches \parencite{sanchez-gonzalez_learning_nodate} may offer a more natural framework for three-dimensional multi-particle configurations, while autoregressive neural emulators \parencite{koehler_apebench_nodate} could address the temporal dimension.

\paragraph{Uncertainty quantification.}
The current framework produces a single deterministic prediction for each input configuration, without any estimate of how reliable or uncertain that prediction is. Adding predictive uncertainty through Bayesian extensions \parencite{zhu_bayesian_2018} or ensembling approaches would increase the practical reliability of the surrogate for geophysical applications, where assessment of prediction confidence is important.
