% ----------------------------------------------------------------------------
% Chapter 11: Discussion
% ----------------------------------------------------------------------------
%
% ============================================================
% QUELLENVORSCHLÄGE FÜR DIESES KAPITEL
% ============================================================
% Architekturvergleich U-Net vs. FNO (Stärken/Schwächen):
%   \parencite{lu_comprehensive_2022}        -- Systematischer Vergleich neural operators
%   \parencite{qin_toward_2024}              -- Spektralbias FNO: warum Fehler nahe Kristallrand
%   \parencite{wen_u-fnoenhanced_2022}       -- U-FNO als mögliche Verbesserung (Ausblick)
%   \parencite{raonic_convolutional_2023}    -- Convolutional Neural Operators als Alternative
%   \parencite{alimehmeti-gopel_weight_2024} -- ELR-Theorie: erklärt U-Net-Instabilität bei hohen lr via BatchNorm-Dynamik
%
% Generalisierungsgrenzen:
%   \parencite{morimoto_generalization_2022} -- Faktoren, die Out-of-Distribution-Performance begrenzen
%   \parencite{rana_scalable_2024}           -- Domain-Decomposition als Skalierungsstrategie
%   \parencite{sanchez-gonzalez_learning_nodate} -- GNN-basierte Alternativen für viele Partikel
%
% Physikalische Einschränkungen / Inductive Bias:
%   \parencite{richter-powell_neural_nodate} -- Divergenzfreie Repräsentationen
%   \parencite{zhu_physics-constrained_2019} -- Physics-Constrained DL als Alternative
%   \parencite{kovachki_neural_nodate}       -- Neural Operator-Theorie (Approximierbarkeit)
%
% Geowissenschaftlicher Kontext:
%   \parencite{martin_crystal_1988}          -- Geophysikalische Relevanz der Ergebnisse
%   \parencite{verhoeven_numerical_2009}     -- Numerische Vergleichsbasis
%   \parencite{patocka_settling_2020}        -- Weitere Referenz für Sedimentation
%
% Zukünftige Arbeiten:
%   \parencite{hu_applying_2023}             -- PINN für Partikelströmung: physik-getriebene Erweiterung
%   \parencite{zhu_bayesian_2018}            -- Unsicherheitsquantifizierung als nächster Schritt
%   \parencite{koehler_apebench_nodate}      -- APEBench: Benchmark für autoregressive Erweiterungen
%   \parencite{li_fourier_nodate}            -- Geo-FNO: Erweiterung auf nicht-rechteckige Domänen
% ============================================================
%
\chapter{Discussion}
\label{ch:discussion}

This chapter interprets the experimental results presented in Chapter~\ref{ch:results}. The central question addressed in this thesis is to what extent U-Net and FNO architectures, trained on configurations with up to a fixed number of crystals, can generalize to unseen sedimentation scenarios with varying particle numbers and spatial arrangements—and which factors fundamentally limit this generalization. Rather than repeating the numerical findings, the aim is to explain why the observed patterns occur and what they imply for the surrogate modeling of multi-crystal Stokes flow.

% ============================================================================
\section{Training Behavior: FNO Stability vs.\ U-Net Oscillations}
% ============================================================================

One of the most consistent findings across all experiments is the difference in training stability between the two architectures. The FNO converges smoothly and monotonically throughout training across all configurations in Experiments 1--3, while the U-Net exhibits pronounced oscillations in the validation loss, particularly at higher learning rates and with the larger batch size. It is worth noting that smooth convergence does not guarantee accurate predictions in every scenario: in Experiment~4, the FNO also converges stably, yet reaches a high error on large crystals due to spectral truncation (discussed further in the crystal-size generalization section below), making this the one setting in which the U-Net outperforms the FNO.

This contrast can be understood in terms of the architectural differences between the two models. The U-Net employs Batch Normalization after each convolutional layer, a design choice that introduces a normalization-scale invariance: the gradient magnitude with respect to any weight is inversely proportional to the current weight norm \parencite{alimehmeti-gopel_weight_2024}. This so-called effective learning rate (ELR) can vary substantially between layers, and at high learning rates the ELR ratios between different layers can temporarily diverge—a phenomenon that is particularly pronounced at the start of training, when weight norms are small and gradients are correspondingly large. The observed oscillations at $\mathrm{lr} = 10^{-3}$ and $5 \times 10^{-3}$ are consistent with this explanation.

By contrast, the FNO does not rely on Batch Normalization in its spectral layers. The gradient clipping applied during FNO training (ClipNorm with threshold $1.0$, see Table~\ref{tab:hyperparams}) provides a direct and architecture-independent mechanism to prevent unstable gradient updates, which largely eliminates the oscillatory behavior seen in the U-Net.

The effect of batch size provides further evidence for this interpretation. For the U-Net, batch size~8 consistently yields lower evaluation errors and smoother training curves than batch size~16. Smaller batches introduce more stochastic gradient noise into the gradient estimates, which effectively reduces the step size in parameter space and can help the optimizer to avoid overshooting. In the language of ELR theory, this corresponds to a reduction of the effective learning rate that keeps the training dynamics in a more stable, subcritical regime. For the FNO, batch size has a smaller effect, consistent with the absence of the BatchNorm-induced ELR dynamics.

% ============================================================================
\section{In-Distribution Accuracy: FNO Advantage in the Single-Crystal Regime}
% ============================================================================

For single-crystal flows (Experiment~1), the best FNO configurations achieve mean relative $L_2$ errors of $1.6$--$2.3\%$ on~$\psi$ and $3.0$--$4.2\%$ on the derived velocity field after 100~training epochs, substantially outperforming the best U-Net configurations, which reach $5.5$--$6.5\%$ on~$\psi$ and $8.5$--$12.7\%$ on velocity after 180~training epochs. In the multi-crystal experiments, where the FNO is trained for 100~epochs and the U-Net for 180~epochs, the FNO still achieves substantially lower errors, confirming that the spectral architecture learns the structure of Stokes flow more efficiently than the U-Net's local convolutional processing.

This performance gap is plausible, given the physical and mathematical structure of the problem. Stokes flow around a single crystal is a globally smooth, elliptic field: the crystal perturbation decays algebraically over distances of several crystal diameters, and there are no sharp shocks or discontinuities in the velocity field outside the crystal boundary itself. The FNO's spectral architecture is structurally well suited for such problems \parencite{li_fourier_2021}: by operating on globally defined Fourier modes, it can capture long-range correlations in a single layer. The U-Net, in contrast, builds global context through a hierarchy of local convolutions and pooling operations, which may require more parameters and training data to accurately represent smooth global fields.

The spectral bias of the FNO—its tendency to represent low-frequency components more accurately than high-frequency ones—is generally a limitation near sharp crystal boundaries \parencite{qin_toward_2024}. However, in the single-crystal regime this limitation appears to have only a moderate effect, since the boundary-layer contribution to the global $L_2$ error is relatively small compared to the domain-wide flow structure. This trade-off is likely to become more important for denser crystal configurations, where the fraction of the domain influenced by boundary layers increases.

A spatial error analysis (Section~\ref{ssec:results:exp1:spatial}) confirms that the FNO generalizes uniformly across the domain: errors of $0.9$–$2.3\%$ show no dependence on crystal position. The U-Net exhibits a mildly positive regression slope of error vs.\ crystal distance from the domain centre, suggesting a weak tendency for larger errors when the crystal is located further from the centre. However, with only ten evaluation samples the statistical significance of this trend is limited, and the dominant factor is the overall higher error level of the U-Net rather than a systematic positional effect.

% ============================================================================
\section{Effect of the Composite Loss on U-Net Accuracy}
% ============================================================================

In Experiment~1, the composite loss function, which augments the standard MSE with gradient-matching and boundary terms, has a clear positive effect on velocity accuracy for batch size~8: the velocity-to-$\psi$ error ratio drops from approximately $3\times$ (plain MSE) to $1.5$--$2\times$ (composite loss), confirming the hypothesis that the elevated velocity errors under plain MSE training originate from high-frequency residuals in~$\hat\psi$ that are amplified by the finite-difference differentiation step (Section~\ref{ssec:results:exp1:unet:hf}).

For batch size~16, however, the composite loss leads to substantially \emph{higher} evaluation errors than plain MSE training (Section~\ref{ssec:results:exp1:unet:metrics}). This is a somewhat unexpected finding. One possible explanation is that the additional gradient and boundary terms modify the effective gradient magnitudes differently across layers, increasing the ELR spread—an effect that would be particularly detrimental for the already less stable bs\,16 training regime. Another contributing factor is that the gradient loss computed via finite differences over the predicted field introduces numerical noise that acts as an additional source of instability when combined with the less regularized gradient estimates of larger batches.

Whatever the precise mechanism, the results indicate that the benefits of the composite loss depend critically on the batch size, and that a simple uniform adoption of additional loss terms does not guarantee improved accuracy.

% ============================================================================
\section{High-Frequency Artifacts: Training Transient vs.\ Structural Limitation}
% ============================================================================

Both architectures produce predictions with visible high-frequency artifacts in parts of the domain, particularly in the derived velocity field. However, the origin and implications of these artifacts differ substantially between the two models, and conflating them would lead to incorrect conclusions about each architecture's suitability.

For the FNO, the high-frequency artifacts are best interpreted as a \emph{training transient} rather than a structural failure. The FNO's spectral layers explicitly truncate the representation at $k_{\max} = 16$ Fourier modes, so frequency components above this cutoff cannot be produced by the spectral convolution path. The high-frequency residuals visible in the predictions therefore originate from the pointwise linear layers $W$ in the FNO architecture, which operate locally in physical space and impose no frequency constraint. Under the spectral bias of neural networks \parencite{qin_toward_2024}—the tendency to fit low-frequency components before high-frequency ones—these pointwise residuals are the last components to converge during training. In Experiments~2 and~3, where the FNO is trained for 100~epochs, the training curves have not fully plateaued, and the high-frequency content in the predictions would be expected to decay with continued training. From this perspective, filtering the predictions post hoc is technically possible and would reduce the apparent error, but it would address a training artifact rather than a structural deficiency of the architecture.

For the U-Net, the situation is qualitatively different. In the single-crystal regime, the composite loss already partially suppresses high-frequency artifacts in $\hat\psi$ via the gradient penalty term, and the remaining residuals are modest relative to the global prediction error. In the multi-crystal regime, however, high-frequency artifacts become more prominent, precisely where the overall prediction quality degrades—that is, in the regions between and near multiple crystal boundaries. This spatial correlation between artifact severity and error magnitude suggests a \emph{structural limitation}: the U-Net must simultaneously represent the global, smoothly varying flow field and the local, crystal-scale boundary effects, using a fixed hierarchical convolutional architecture. As the number of crystals increases, these two representational demands compete for model capacity, and the U-Net increasingly resolves the tension by approximating the global structure while producing incoherent high-frequency outputs near the local boundary regions. In this case, a post hoc filter would remove the visual artifacts but would not reduce the underlying prediction error, since the latter reflects genuine inaccuracies in the predicted $\psi$ values rather than superimposed noise.

In summary, the high-frequency artifacts of the FNO are primarily a consequence of insufficient training time and would diminish with further optimization, whereas the corresponding artifacts in the U-Net for multi-crystal configurations indicate an architectural capacity constraint.

% ============================================================================
\section{Multi-Crystal Generalization and Its Limits}
% ============================================================================

Both architectures show substantially higher errors in the multi-crystal regime compared to the single-crystal baseline, even within the training distribution. In contrast to the U-Net, the FNO converges steadily on the multi-crystal task: both training MSE and validation error decrease monotonically throughout the 100~training epochs, indicating that the chosen hyperparameters remain well-suited as the geometric complexity increases. A notable observation is that the U-Net's best checkpoint in Experiment~2 is reached already at epoch~43 (validation relative $L_2$ error of $28.5\%$), with no further improvement over the remaining $137$ epochs. This suggests that the U-Net's best representable solution for the multi-crystal task is largely found in the early phase of training, with the subsequent training time spent on oscillations around that solution rather than genuine refinement. The near-identical best-checkpoint error in Experiment~3 ($28.6\%$ at epoch~49) compared to Experiment~2 ($28.5\%$ at epoch~43) further suggests that the wider training distribution ($n \leq 25$ instead of $n \leq 10$) provides no significant benefit for U-Net convergence behavior. The FNO reaches $4$--$9\%$ on~$\psi$ for $n \leq 10$ (Experiment~2), while the U-Net reaches $17$--$26\%$. In absolute terms, the degradation relative to the single-crystal baseline is considerably larger for the U-Net—rising by roughly 11--21~percentage points from its Experiment~1 best of ${\sim}5.5\%$—than for the FNO, which increases by only 2--7~percentage points from its Experiment~1 best of ${\sim}1.6\%$, confirming that the spectral architecture generalizes more efficiently across geometric complexity.

A physically motivated explanation for the U-Net's relative weakness in the multi-crystal regime is related to the superposition principle of Stokes flow \parencite{happel_low_1983}. For multiple crystals, the flow field is approximately the sum of the individual crystal contributions (modulated by hydrodynamic interactions). Representing such superpositions accurately requires the network to capture long-range correlations between distant crystal positions—a task for which the FNO's global spectral convolutions are better suited than the hierarchical local operations of the U-Net. The elevated velocity errors relative to $\psi$ errors even within the training distribution (Experiment~3: $27$–$38\%$ vs.\ $19$–$29\%$) confirm that the U-Net struggles with the superposition of flow contributions already before any distribution shift occurs.

At very high crystal counts ($n = 67$–$84$), the wide per-sample spread in errors ($\psi$: $18$–$32\%$, velocity: $33$–$45\%$) reflects high sensitivity to the specific crystal arrangement: some configurations happen to be closer to those seen during training and are predicted more accurately, while others are not. This variability is consistent with the U-Net's reliance on local spatial patterns—whether a dense arrangement locally resembles something in the training set is essentially a matter of chance at extreme crystal counts.

Beyond the training boundary, the degradation patterns of the two architectures diverge markedly. In Experiment~2, the FNO shows a gradual performance increase to $10.1\%$ at $n = 20$ and $13.8\%$ at $n = 25$ (compared to $\leq 9.2\%$ within the training distribution ($n \leq 10$)), while the U-Net maintains a roughly constant error level regardless of crystal count. This pattern is consistent with a distribution shift interpretation: the FNO has learned representations that are tightly coupled to the density regime of the training distribution, while the U-Net, being already at a high error level within the distribution, does not degrade further in the OOD regime—a case of graceful degradation that is less a sign of robustness than of uniformly poor performance. Notably, the U-Net's near-OOD generalization ($n \in \{11, 12, 15\}$) is broadly consistent with in-distribution performance, suggesting smooth extrapolation to slightly higher crystal numbers. At $n \in \{20, 25\}$, however, a velocity-specific degradation emerges: the velocity error rises to ${\sim}31$–$35\%$ while the $\psi$ error remains comparable (${\sim}22$–$27\%$), indicating that the derived flow field degrades before the stream function itself does—consistent with the U-Net producing increasingly incoherent high-frequency content in the velocity-relevant gradients of~$\hat\psi$.

Experiment~3 confirms that the FNO's OOD degradation is primarily driven by distribution shift rather than architectural limitations: training on a wider distribution ($n \leq 25$) substantially extends the generalization range, allowing the FNO to maintain errors below $10\%$ up to $n = 35$—considerably further than the $n = 20$ threshold at which Experiment~2 first exceeds $10\%$ (just twice the training maximum of $n = 10$). This finding is in line with \textcite{morimoto_generalization_2022}, who showed that the generalization range of neural surrogates for multi-obstacle flows scales approximately with the training distribution width.

Across Experiments~2 and~3, the FNO consistently outperforms the U-Net in-distribution by a factor of roughly $3$–$4\times$. In the OOD regime, the advantage narrows but persists: at $n = 50$, the FNO reaches $12.4\%$ on~$\psi$ and ${\sim}14\%$ on velocity, compared to $21.5\%$ and ${\sim}33\%$ for the U-Net — a remaining factor of roughly $1.7\times$ on~$\psi$ and $2.4\times$ on velocity. The consistent gap across all three experiments, which widens with increasing geometric complexity, indicates that the FNO's advantage is not limited to the simplest configurations but scales with problem difficulty.

% ============================================================================
\section{Physical Consistency: The Stream-Function Representation}
% ============================================================================

Predicting the stream function $\psi$ rather than the velocity components directly ensures that incompressibility is satisfied analytically by construction: since the velocity recovered via Eq.~\ref{eq:stream_function_def} is divergence-free for any differentiable $\psi$ \parencite{richter-powell_neural_2022}, no penalty term for mass conservation is required during training. The divergence RMS values of all FNO predictions are on the order of $10^{-26}$, confirming that this analytical guarantee holds exactly in practice.

This is a significant advantage over direct velocity prediction, where small but systematic divergence errors can accumulate and lead to non-physical flow structures. It is also more efficient than enforcing incompressibility through a soft constraint term in the loss function \parencite{zhu_physics-constrained_2019}, as no additional hyperparameter tuning is required.

The one downside of the stream-function representation is that the velocity field must be recovered through finite differences, which amplifies high-frequency residuals in $\hat\psi$ into larger relative errors in~$\mathbf{u}$. This effect is more pronounced for the U-Net than for the FNO, as shown in Sections~\ref{ssec:results:exp1:unet:hf} and~\ref{ssec:results:exp1:comparison}. The composite loss partially mitigates this issue, but a more direct solution would be to train on the velocity field directly—at the cost of losing the incompressibility guarantee.

% ============================================================================
\section{Crystal-Size Generalization: A Reversal of the Architectural Ranking}
% ============================================================================

Experiment~4 introduces an additional axis of out-of-distribution generalization: both models are evaluated on single-crystal configurations with crystal radii outside the range used during training—either smaller (``small'' crystals) or larger (``large'' crystals) than any training example. Qualitative results (Section~\ref{ssec:results:exp4:qualitative}) are shown for the FNO only, as it exhibits the more striking failure mode for large crystals and the trained-vs-untrained comparison is most informative there. The untrained baseline confirms that the predicted flow structure is genuinely learned: without any trained weights the output is essentially featureless, so all structure visible in the trained predictions originates from the learned mapping.

For small crystals, the two architectures perform comparably, both achieving relative $L_2$ errors of approximately $11$--$12\%$ on~$\psi$ and $20\%$ on the derived velocity field. The relatively large standard deviations ($\pm 6.4\%$ for the FNO, $\pm 6.8\%$ for the U-Net) indicate that accuracy varies considerably across individual samples, suggesting that the specific crystal placement interacts with the OOD size in ways that are not uniform. This is somewhat elevated relative to the in-distribution errors of Experiment~1, but not dramatically so: small crystals produce weaker, more diffuse flow perturbations, which the trained models can partially represent by drawing on the general structure of Stokes flow they have internalized.

For large crystals, the picture changes substantially. The FNO error reaches $71.3 \pm 4.8\%$ on~$\psi$ and $67.7 \pm 3.9\%$ on velocity—a severe degradation that renders the predictions essentially unreliable. The U-Net, by contrast, achieves $44.9 \pm 8.2\%$ on~$\psi$ and $50.7 \pm 4.4\%$ on velocity. While these numbers are still far from the in-distribution performance, the U-Net's relative advantage over the FNO is reversed compared to Experiments~1--3.

A physically motivated explanation for this reversal is the spectral bias of the FNO. The FNO's Fourier layers are truncated at $k_{\max} = 16$ modes, which fixes the spatial frequencies that can be represented by the spectral convolution. For in-distribution crystal sizes, this truncation is sufficient to capture the boundary-layer structure near the crystal surface. For larger crystals, the boundary layer and the wake region occupy a proportionally larger fraction of the domain, and the sharp transition between the solid and fluid regions introduces higher effective spatial frequencies relative to the domain size. The fixed mode truncation at $k_{\max} = 16$ is insufficient to represent these features accurately, leading to the large errors observed.

The U-Net, whose local convolutional filters operate at fixed spatial scales on the grid, is less sensitive to this type of geometric scaling: the local neighborhood structure around the crystal boundary remains geometrically similar regardless of crystal size, and the U-Net can partially exploit this self-similarity. This structural advantage over the spectrally-truncated FNO is consistent with the findings of \textcite{qin_toward_2024}, who showed that FNOs underperform near sharp interfaces precisely because of the spectral truncation.

These results suggest that, while the FNO is the superior architecture for in-distribution and moderate OOD generalization across crystal counts, the U-Net may be better suited for scenarios that require robustness to geometric scaling.

The variable-radius FNO experiment (Section~\ref{ssec:results:exp4:varradius}) demonstrates that explicit size conditioning substantially mitigates this limitation for the FNO. By training on three discrete radii ($r \in \{0.02, 0.05, 0.08\}$) and leveraging the radius-of-nearest-crystal input channel, the model achieves a mean relative $L_2$ error of $10.4\%$ on~$\psi$ across OOD evaluation sizes—a dramatic improvement over the fixed-radius FNO's $71.3\%$ error for large crystals (Table~\ref{tab:fno_metrics_exp4_var}). This result confirms that the spectral truncation limitation is not a fundamental architectural barrier: providing the network with explicit information about the geometric scale allows it to adapt its effective representation without requiring architectural modifications such as adaptive Fourier modes.

An architecturally more general alternative would be the use of adaptive Fourier modes or multi-resolution spectral architectures that adjust their frequency representation to the geometric scale automatically—without requiring explicit size information in the input channel. Such approaches represent a promising direction for future work, as they would extend the generalization benefit beyond the discrete set of training radii.

A complementary perspective on the FNO's spectral behavior is provided by the frequency-domain analysis of residuals in Section~\ref{sec:results:spectral}. Contrary to the expectation that FNO errors are concentrated beyond the truncation boundary at $k = 16$, the relative error fraction $r(k)$ remains approximately constant and low across all frequencies. This suggests that the primary source of prediction error is not the spectral truncation mechanism itself, but rather the overall model capacity and the difficulty of the regression task in the low-to-medium frequency regime where nearly all signal energy resides—consistent with the observation of \textcite{qin_toward_2024} that FNO errors near sharp interfaces manifest as smooth, low-frequency residuals rather than high-frequency artefacts. In practical terms, increasing $k_{\max}$ beyond $16$ is unlikely to yield substantial accuracy gains for the smooth Stokes flow fields considered here; improvements are more likely to come from increased model width, depth, or training data.

% ============================================================================
\section{Limitations}
% ============================================================================

Several limitations of this study should be noted when interpreting the results.

\paragraph{Training data coverage and interpolation regime.}
A structural concern in the experimental design is the relationship between the size of the training set and the effective coverage of the configuration space. For Experiments~1--3, each crystal-count stratum is represented by 1\,000 training samples with a fixed crystal radius $r = 0.05$ in a $[-1,1]^2$ domain. For $n = 1$, the configuration space is essentially two-dimensional (crystal center position), and 1\,000 samples may cover it densely enough that the held-out test samples are effectively interpolations within the training distribution rather than genuinely unseen configurations. In this regime, the reported test errors may overestimate the model's true generalization capability.

For higher crystal counts, this concern is mitigated in one sense—the configuration space grows combinatorially with $n$—but a competing effect emerges: geometrically valid configurations (without crystal overlap) become rarer, because placing many fixed-radius circles in a bounded domain without mutual overlap grows increasingly constrained—only spatially dispersed arrangements tend to be accepted. The packing constraint thus implicitly biases both training and test distributions toward specific structural arrangements. If the test samples are drawn from the same constrained distribution, they may resemble training samples in a way that artificially improves the reported generalization metrics. A more conservative evaluation protocol would either increase the number of test samples substantially, or explicitly select test configurations that maximize distance to the training set in some geometric measure.

\paragraph{FNO results for Experiments 2 and 3.} The FNO results for Experiments~2 and~3 are based on training runs of 100~epochs, compared to 180~epochs for the U-Net. While the FNO training curves suggest that convergence is well underway at 100~epochs, the reported numbers may improve slightly with extended training. Direct architectural comparisons in Experiments~2 and~3 should therefore be interpreted with this asymmetry in mind.

\paragraph{Two-dimensional and steady-state assumption.} All simulations and predictions are restricted to steady-state, two-dimensional Stokes flow with circular crystals. Real magmatic sedimentation is inherently three-dimensional and time-dependent. The relevance of the current results for three-dimensional transient flows remains to be demonstrated.

\paragraph{Crystal size generalization.} Experiment~4 demonstrates that both architectures degrade significantly for crystals larger than the training distribution, with the FNO more severely affected than the U-Net. The evaluation is based on only ten samples per size category, and the models used were not specifically trained for size robustness. A dedicated training strategy incorporating variable crystal sizes would be needed to achieve reliable size generalization.

